---
title: "Assignment 2.1: School composition and school choice"
author: "Mariia Steeghs-Turchina · DV7"
output:
  html_document:
    toc: true
    self_contained: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE,
    fig.width = 8, fig.height = 4.5, fig.align = "center", out.width = "100%"
)
```

## Research question

Among regular Dutch primary schools, is the neighbourhood composition of
four-year-old intake associated with that of pupils already enrolled in
the previous year, after accounting for where families live?

**Level: inference.** If families respond to a school's existing composition,
a small change in intake could persist. We ask whether the data are
consistent with that possibility. They cannot by themselves show that
composition causes school choice.

**Target audience.** Education officers concerned with segregation and
parents choosing schools. The visualization should connect
school-level patterns to those decisions.

## Datasets

We use four public datasets:

- **[DUO pupil origins (2022–2023)](https://onderwijsdata.duo.nl/dataset/4ea91ae5-51ef-48a4-aa67-7098343f75bc):**
  counts by school location and home postcode (PC4).
  We use these to calculate each school's neighbourhood composition.
- **[DUO school totals (2022–2023)](https://onderwijsdata.duo.nl/dataset/cf80e90d-ed19-4a10-a138-00c5c345cb5e):**
  total enrolment per school location.
  We use these as the denominator for coverage.
- **[CBS postcode population (2022)](https://opendata.cbs.nl/#/CBS/nl/dataset/83503NED/table):**
  residents by migration background and PC4.
  We use these to describe the neighbourhoods linked to pupils' postcodes.
- **[DUO pupils by age and postcode (2023)](https://duo.nl/open_onderwijsdata/images/03.-leerlingen-po-totaaloverzicht-2023-2024.csv):**
  counts by age, school, and home postcode.
  We inspect age-four cells as a possible intake measure and quantify suppression.

## Exploratory process

### Data and cleaning

The download script caches the source files.
Functions in `R/` then prepare, inspect, and plot the data.

```{r data}
library(ggplot2)
library(plotly)
library(scales)

source("scripts/download.R")
source("R/prepare.R")
source("R/plots.R")

data <- prepare_data(paths)
intake <- check_intake(paths[["intake"]])
schools_2023 <- filter(data$schools, year == 2023)
selected <- filter(schools_2023, eligible)
#' Format counts with thousands separators.
format_n <- function(x) format(x, big.mark = ",", scientific = FALSE)
```

Each row in the DUO origins data represents one school–postcode–year; each
plotted point represents one school. We keep school codes as text, retain
regular primary schools in 2022 and 2023, and join CBS to pupils' home PC4.

The **neighbourhood composition proxy** is the pupil-weighted average of
the share of residents in CBS's historical “non-western migration
background” category. Both enrolment years use the 2022 CBS release.

DUO marks suppressed counts as `-1`, we treat them as missing. A postcode
enters the weighted average only when its count is exact and CBS reports at
least 50 residents. A school is eligible when at least 50 pupils are matched
and coverage is at least 80%.

```{r coverage}
plot_coverage(data$schools)
```

Of `r format_n(nrow(schools_2023))` regular primary school locations in 2023,
`r format_n(sum(is.na(schools_2023$coverage)))` lack a usable total and
`r format_n(nrow(selected))` pass both quality rules.
Overall median coverage is
`r percent(median(schools_2023$coverage, na.rm = TRUE), accuracy = 0.1)`.
Coverage combines count suppression, unusable postcodes, and missing CBS
values.

### First patterns

```{r composition}
plot_composition(data$schools)
```

The middle half of included schools spans
`r percent(quantile(selected$composition, 0.25), accuracy = 0.1)` to
`r percent(quantile(selected$composition, 0.75), accuracy = 0.1)`.
This variation makes the question worth investigating, but may reflect
residential segregation and geography as well as school choice.

```{r persistence-interactive}
ggplotly(plot_persistence(data$pairs), tooltip = "text")
```

Among `r format_n(nrow(data$pairs))` schools meeting the rules in both years,
the median absolute change is
`r round(100 * median(abs(data$pairs$change)), 2)` percentage points.
The `r sum(abs(data$pairs$change) > 0.05)` orange triangles flag changes
larger than five points for closer inspection.
Large changes could also arise from mergers or changing data coverage.

### Next stage

We also checked DUO's October 2023 file by age and home postcode. Among
`r format_n(intake$positive_cells)` school–postcode cells with at least one
four-year-old, `r percent(intake$suppressed_share, accuracy = 0.1)`
are suppressed as `<5`. Age four is a proxy for intake.
Small flows will therefore need a defensible treatment before modelling.

- Keep neighbourhood composition as a labelled proxy.
- Examine four-year-old intake separately from existing pupils. Most
  pupils remain enrolled between years, so persistence in total enrolment
  is expected.
- Check school continuity and coverage before interpreting large changes.
- For the later inference, account for home-postcode and school differences
  and existing recruitment links. Show estimated associations with
  uncertainty; these exploratory plots do not answer the causal question.

### Sources and reproducibility

[DUO pupil origins](https://onderwijsdata.duo.nl/dataset/4ea91ae5-51ef-48a4-aa67-7098343f75bc),
[DUO school totals](https://onderwijsdata.duo.nl/dataset/cf80e90d-ed19-4a10-a138-00c5c345cb5e),
[DUO pupils by age and postcode](https://duo.nl/open_onderwijsdata/images/03.-leerlingen-po-totaaloverzicht-2023-2024.csv),
and [CBS table 83503NED](https://opendata.cbs.nl/#/CBS/nl/dataset/83503NED/table).
The first knit downloads these public sources; later knits reuse local
copies. The raw data is not committed. Source releases may be revised.
Code: [thatmariia/bdst_dv](https://github.com/thatmariia/bdst_dv).
