---
title: "Assignment 2.1: School composition and school choice"
author: "Mariia Steeghs-Turchina · DV7"
output:
  html_document:
    toc: true
    self_contained: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE,
    fig.width = 8, fig.height = 4.5, fig.align = "center", out.width = "100%"
)
```

## Research question

Among regular Dutch primary schools, is the neighbourhood composition of
four-year-old intake associated with that of pupils already enrolled in
the previous year, after accounting for where families live?

**Level: inference.** If families consider who already attends a school,
small differences in pupil composition could reinforce themselves over
time. This question explores whether school choice might contribute to
lasting segregation beyond differences in where families live.

## Exploratory process

### Data and cleaning

The download script caches the source files. Functions in `R/` then prepare,
inspect, and plot the data.

```{r data}
library(ggplot2)
library(scales)

source("scripts/download.R")
source("R/prepare.R")
source("R/plots.R")

data <- prepare_data(paths)
intake <- check_intake(paths[["intake"]])
schools_2023 <- filter(data$schools, year == 2023)
selected <- filter(schools_2023, eligible)
#' Format counts with thousands separators.
format_n <- function(x) format(x, big.mark = ",", scientific = FALSE)
```

We use DUO enrolment by school location and home postcode (PC4), DUO school
totals, and CBS population by PC4. One raw DUO row is a school–postcode–year;
one plotted observation is a school location. School codes stay as text to
preserve leading zeros. We restrict exploration to regular primary schools
in 2022 and 2023 and join CBS by the home postcode, never the school's address.

The **neighbourhood composition proxy** is the pupil-weighted average of
the share of residents in CBS's historical “non-western migration
background” category in pupils' home postcodes. It describes neighbourhoods,
**not individual pupils' backgrounds**. Both enrolment years use CBS 2022
so that postcode population changes do not drive the comparison.

Suppressed DUO counts (`-1`) remain missing. Only exact counts with a valid
CBS match and at least 50 postcode residents enter the weighted average.
We require at least 50 matched pupils per school and coverage of at least
80% of its total enrolment. These are exploratory quality rules; omitted
pupils may differ from those observed.

```{r coverage}
plot_coverage(data$schools)
```

Of `r format_n(nrow(schools_2023))` regular primary school locations in 2023,
`r format_n(sum(is.na(schools_2023$coverage)))` lack a usable total and
`r format_n(nrow(selected))` pass both quality rules.
Overall median coverage is
`r percent(median(schools_2023$coverage, na.rm = TRUE), accuracy = 0.1)`.
Coverage combines count suppression, unusable postcodes, and missing CBS
values; it is not a pure postcode match rate.

### First patterns

```{r composition}
plot_composition(data$schools)
```

The middle half of included schools spans
`r percent(quantile(selected$composition, 0.25), accuracy = 0.1)` to
`r percent(quantile(selected$composition, 0.75), accuracy = 0.1)`.
This variation makes the question worth investigating, but may reflect
residential segregation and geography as well as school choice.

```{r persistence, fig.height = 6.5}
plot_persistence(data$pairs)
```

Among `r format_n(nrow(data$pairs))` schools meeting the rules in both years,
the median absolute change is
`r round(100 * median(abs(data$pairs$change)), 2)` percentage points.
The `r sum(abs(data$pairs$change) > 0.05)` orange triangles flag changes
larger than five points for closer inspection, not automatic exclusion.
Large changes could also arise from mergers or changing data coverage.

### Decisions for the next stage

We also checked DUO's October 2023 file by age and home postcode. Among
`r format_n(intake$positive_cells)` school–postcode cells with at least one
four-year-old, `r percent(intake$suppressed_share, accuracy = 0.1)`
are suppressed as `<5`. This is the share of **cells**, not pupils.
Age four is a proxy for intake, not an exact record of new enrolments.
Small flows will therefore need a defensible treatment before modelling.

- Keep neighbourhood composition as an explicitly labelled proxy; avoid
  claims about individual families or a school's quality.
- Examine four-year-old intake separately from existing pupils. Most
  pupils remain enrolled between years, so persistence in total enrolment
  is expected and does not demonstrate feedback.
- Check school continuity and coverage before interpreting large changes;
  compare results under stricter coverage rules.
- For the later inference, account for home-postcode and school differences
  and existing recruitment links. Show estimated associations with
  uncertainty; these exploratory plots do not answer the causal question.

### Sources and reproducibility

[DUO pupil origins](https://onderwijsdata.duo.nl/dataset/4ea91ae5-51ef-48a4-aa67-7098343f75bc),
[DUO school totals](https://onderwijsdata.duo.nl/dataset/cf80e90d-ed19-4a10-a138-00c5c345cb5e),
[DUO pupils by age and postcode](https://duo.nl/open_onderwijsdata/images/03.-leerlingen-po-totaaloverzicht-2023-2024.csv),
and [CBS table 83503NED](https://opendata.cbs.nl/#/CBS/nl/dataset/83503NED/table).
The first knit downloads these public sources; later knits reuse local
copies. No raw data is committed. Source releases may be revised.
Code: [thatmariia/bdst_dv](https://github.com/thatmariia/bdst_dv).
